---
title: "St. Louis MSA Historic Population Change"
author: "Christopher Prener, Ph.D"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates a county-level historic data set for the St. Louis, MO.

## Dependencies
This notebook requires a range of tidyverse and spatial packages:

```{r load-packages}
# tidyverse packages
library(dplyr)
library(readr)
library(stringr)
library(tidyr)

# spatial packages
library(measurements)
library(tidycensus)
library(tigris)
library(sf)

# other packages
library(here)
```

## Load Data
This notebook requires historical data accessed from the [U.S. Census Bureau](https://www.census.gov/population/www/censusdata/pop1790-1990.html). These data have been manually cleaned to isolate counties of interest from Missouri and Illinois, given the poor format of the original spreadsheet.

```{r load-data}
historic <- read_csv(here("data", "historicPopulations.csv"))
```

## Convert to Long Form
The data were cleaned manually in as minimal fashion as possible, so there is still additional cleaning to be done. We need to convert them to long, change the county name to something distinct for pre-divorce St. Louis, and re-order the data:

```{r clean-historic}
historic %>%
  gather(key = "year", value = "pop", -State, -County) %>%
  mutate(County = ifelse(County == "St. Louis" & year <= 1870, "St. Louis orig", County)) %>%
  arrange(State, County, year) %>%
  filter(is.na(pop) == FALSE) %>%
  rename(
    county = County,
    statefp = State
  ) %>%
  mutate(year = as.numeric(year)) -> historicLong
```

## Download Contemporary Census Data
Next, we'll want to add data from 2000, 2010, and 2017 (the most recent ACS year available) to our data set.

### 2000
We need to download the Missouri and Illinois data, do some minor cleaning, and then combine them.

```{r census-2000}
# create vector of missouri counties
counties <- c(29071, 29099, 29113, 29183, 29189, 29219, 29510)

# download county populations
moCounties <- get_decennial(geography = "county", variables = "P001001", year = 2000, state = 29) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 29)

# create vector of illinois counties
counties <- c(17027, 17083, 17119, 17133, 17163)

# download county populations
ilCounties <- get_decennial(geography = "county", variables = "P001001", year = 2000, state = 17) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 17)

# combine and clean
bind_rows(moCounties, ilCounties)%>%
  select(-variable) %>%
  rename(
    county = NAME,
    pop = value
  ) %>%
  mutate(county = str_trim(str_replace(string = county, pattern = "County", replacement = ""))) %>%
  mutate(year = 2000) %>%
  select(GEOID, county, statefp, year, pop) -> census2000
```

### 2010
We'll repeat the process for 2010:

```{r census-2010}
# create vector of missouri counties
counties <- c(29071, 29099, 29113, 29183, 29189, 29219, 29510)

# download county populations
moCounties <- get_decennial(geography = "county", variables = "P001001", year = 2010, state = 29) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 29) %>%
  mutate(NAME = str_trim(str_replace(string = NAME, pattern = ", Missouri", replacement = "")))

# create vector of illinois counties
counties <- c(17027, 17083, 17119, 17133, 17163)

# download county populations
ilCounties <- get_decennial(geography = "county", variables = "P001001", year = 2010, state = 17) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 17) %>%
  mutate(NAME = str_trim(str_replace(string = NAME, pattern = ", Illinois", replacement = "")))

# combine and clean
bind_rows(moCounties, ilCounties)%>%
  select(-variable) %>%
  rename(
    county = NAME,
    pop = value
  ) %>%
  mutate(county = str_trim(str_replace(string = county, pattern = "County", replacement = ""))) %>%
  mutate(year = 2010) %>%
  select(GEOID, county, statefp, year, pop) -> census2010
```

### 2017
The process is very similar for 2017, except that these are ACS data so the variables requested and returned are different:

```{r acs-2017}
# create vector of missouri counties
counties <- c(29071, 29099, 29113, 29183, 29189, 29219, 29510)

# download county populations
moCounties <- get_acs(geography = "county", variables = "B01003_001", year = 2017, state = 29) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 29) %>%
  mutate(NAME = str_trim(str_replace(string = NAME, pattern = ", Missouri", replacement = "")))

# create vector of illinois counties
counties <- c(17027, 17083, 17119, 17133, 17163)

# download county populations
ilCounties <- get_acs(geography = "county", variables = "B01003_001", year = 2017, state = 17) %>%
  filter(GEOID %in% counties) %>%
  mutate(statefp = 17) %>%
  mutate(NAME = str_trim(str_replace(string = NAME, pattern = ", Illinois", replacement = "")))

# combine and clean
bind_rows(moCounties, ilCounties)%>%
  select(-variable, -moe) %>%
  rename(
    county = NAME,
    pop = estimate,
  ) %>%
  mutate(county = str_trim(str_replace(string = county, pattern = "County", replacement = ""))) %>%
  mutate(year = 2017) %>%
  select(GEOID, county, statefp, year, pop) -> acs2017
```

### Combine
Next, we'll combine them and do some additional data cleaning to make our analytic data set:

```{r combine-data}
bind_rows(historicLong, census2000, census2010, acs2017) %>%
  arrange(statefp, county, year) %>%
  mutate(GEOID = case_when(
    county == "Clinton" ~ "17027",
    county == "Jersey" ~ "17083",
    county == "Madison" ~ "17119",
    county == "Monroe" ~ "17133",
    county == "St. Clair" ~ "17163",
    county == "Franklin" ~ "29071",
    county == "Jefferson" ~ "29099",
    county == "Lincoln" ~ "29113",
    county == "St. Charles" ~ "29183",
    county == "St. Louis" ~ "29189",
    county == "St. Louis city" ~ "29510",
    county == "Warren" ~ "29219"
  )) %>%
  mutate(countyfp = case_when(
    county == "Clinton" ~ 27,
    county == "Jersey" ~ 83,
    county == "Madison" ~ 119,
    county == "Monroe" ~ 133,
    county == "St. Clair" ~ 163,
    county == "Franklin" ~ 71,
    county == "Jefferson" ~ 99,
    county == "Lincoln" ~ 113,
    county == "St. Charles" ~ 183,
    county == "St. Louis" ~ 189,
    county == "St. Louis city" ~ 510,
    county == "Warren" ~ 219
  )) %>%
  select(GEOID, county, statefp, countyfp, year, pop) -> pop
```

## Clean-up Enviornment
Before we move on, we'll drop unneeded objects:

```{r clean-enviornment-1}
rm(acs2017, census2000, census2010, historic, historicLong, ilCounties, moCounties, counties)
```

## Write Data
Next, we'll write these data to `.csv`:

```{r write-pop-data}
write_csv(pop, here("data", "populationData.csv"))
```

