---
title: "Population Trends"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates an initial series of plots for population changes in the St. Louis metro area.

## Dependencies
This notebook requires a number of `tidyverse` packages:

```{r load-packages}
# tidyverse packages
library(dplyr)
library(ggplot2)
library(readr)

# other packages
library(here)
library(prener)
library(RColorBrewer)
library(scales)
```

## Load Data
This notebook requires the previously created population data for the St. Louis MSA (see `createData.Rmd`):

```{r load-data}
pop <- read_csv(here("data", "populationData.csv"))
```

## St. Louis Population
First, we'll make a plot for just the City of St. Louis and St. Louis County:

```{r plot-city-county}
pop %>%
  filter(county == "St. Louis" | county == "St. Louis city" | county == "St. Louis orig") %>%
  mutate(county = ifelse(county == "St. Louis orig", "St. Louis, pre-1876", county)) %>%
  ggplot(mapping = aes(year, pop, by = county)) +
    geom_line(aes(color = county), size = 2) +
    scale_x_continuous(breaks = c(1800, 1820, 1840, 1860, 1880, 1900, 1920, 1940, 1960, 1980, 2000, 2017)) +
    scale_colour_brewer("County", palette = "Set1") +
    labs(
      title = "St. Louis Historic Population Change",
      subtitle = "Before and After the Great Divorce (1876)",
      caption = "Plot by Christopher Prener, Ph.D.\nData via U.S. Census Bureau",
      x = "Year",
      y = "Population"
    ) +
  cp_sequoiaTheme(background = "gray", base_size = 23) -> plot

# save plot
cp_plotSave(filename = here("results", "historicChange.png"), plot = plot, preset = "med")

# print plot
plot
```

Next, we'll plot out population change in the Missouri Counties outside of St. Louis City and County but in the MSA:

```{r plot-mo}
pop %>%
  filter(year >= 1940) %>%
  filter(statefp == 29) %>%
  filter(county != "St. Louis" & county != "St. Louis city") %>%
  ggplot(mapping = aes(year, pop, by = county)) +
    geom_line(aes(color = county), size = 2) +
    scale_y_continuous(labels = comma) + 
    scale_x_continuous(breaks = c(1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
    scale_colour_brewer("County", palette = "Set1") +
    labs(
      title = "Suburban and Exurban Growth, 1940-present",
      subtitle = "St. Louis MSA Counties (Missouri only)",
      caption = "Plot by Christopher Prener, Ph.D.\nData via U.S. Census Bureau",
      x = "Year",
      y = "Population"
    ) +
  cp_sequoiaTheme(background = "gray", base_size = 23) -> plot

# save plot
cp_plotSave(filename = here("results", "suburbanMissouri.png"), plot = plot, preset = "med")

# print plot
plot
```

Finally, we'll plot out population change in the Illinois Counties in the MSA:

```{r plot-illinois}
pop %>%
  filter(year >= 1940) %>%
  filter(statefp == 17) %>%
  ggplot(mapping = aes(year, pop, by = county)) +
    geom_line(aes(color = county), size = 2) +
    expand_limits(y = 300000) + 
    scale_y_continuous(labels = comma, breaks = c(0, 100000, 200000, 300000)) + 
    scale_x_continuous(breaks = c(1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
    scale_colour_brewer("County", palette = "Set1") +
    labs(
      title = "Suburban and Exurban Growth, 1940-present",
      subtitle = "St. Louis MSA Counties (Illinois only)",
      caption = "Plot by Christopher Prener, Ph.D.\nData via U.S. Census Bureau",
      x = "Year",
      y = "Population"
    ) +
  cp_sequoiaTheme(background = "gray", base_size = 23) -> plot

# save plot
cp_plotSave(filename = here("results", "suburbanIllinois.png"), plot = plot, preset = "med")

# print plot
plot
```

